`traveltime`: an R package to calculate walking and motorised travel time from user locations

### 2025-01-24 paper/paper.qmd
several things: - headings resolving #40 - backticks attempting to address #33 - splitting chunks with better chunk options to resolve #38

```{r ttexample,  echo=FALSE}
#| label: "fig-tt_healthcare"
#| fig-cap: "Motorised travel time from healthcare facilities in Cambodia"
library(terra)
library(malariaAtlas)
library(tidyterra)

@@

tt_kh <- getRaster(
  dataset_id = "Accessibility__202001_Global_Motorized_Travel_Time_to_Healthcare",
  shp = boundary_kh
)

@@

Although this article is intended to be the key reference for the `traveltime` package, any citations of it should also be citing the underlying work by `@weiss2018global; @weiss2020global`.

@@

Global maps of travel time to cities `[@weiss2018global]` and health care facilities `[@weiss2020global]` have generated significant interest and use. Collectively `@weiss2018global ; @weiss2020global` have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package `[@geodata]`. There is clear demand for these type of products.

@@

Other R packages produce superficially similar though not identical functionality. At least three R packages provide interfaces to the <https://www.TravelTime.com> API `[@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]`. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location. The latter is most comparable to what the `traveltime` package achieves,

THIS NEXT SENTENCE SUX - EXPLAIN WHY ITS THE SAME AND GIVE A BETTER EXPLANATION OF THE DIFFERENCE.

however without complete spatial coverage, or distance to the nearest of a given point. Furthermore, it requires access keys, caps queries, and uses an external server for calculations, which add friction to the use experience.

@@

# Example workflow: walking in Singapore

@@

To do this we will need two items of information: - our are of interest - in this case, a map of Singapore - the locations of MRT and LRT stations

@@

```{r get basemap}
#| label: "fig-basemap"
#| fig-cap: "Base map of Singapore"
library(terra)
library(geodata)
library(tidyterra)

@@

ggplot() +
  geom_spatvector(
    data = sin
  ) +
  # scale_fill_viridis_c(
  #   option = "A",
  #   direction = -1,
  #   na.value = "transparent"
  # ) +
  scale_fill_manual(na.value = "transparent") +
  theme_minimal()

@@

We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`. It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.

@@

```{r  download friction surface}
library(traveltime)

friction_singapore <- get_friction_surface(
    surface = "walk2020",
    extent = sin
  )|> 
  mask(sin)

friction_singapore
```

Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points. Higher values of friction indicate more time travelling across a given cell.

```{r plot data}
#| label: "fig-data"
#| fig-cap: "Friction surface raster of Singapore, showing boundary from @fig-basemap and station locations as grey points"

@@

@@

Here are those results as in a map:

@@

Note that the above raster include infinite (`Inf`) values. In @fig-data, a number of islands to the south are shown as cells unconnected with the mainland. These raster cells for these islands appear absent in @fig-result. Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot @fig-result.

>>>>

```{r ttexample,  include=FALSE}
library(terra)
library(malariaAtlas)
library(tidyterra)

@@

tt_kh <- getRaster(
  dataset_id = "Accessibility__202001_Global_Motorized_Travel_Time_to_Healthcare",
  shp = boundary_kh
)
```

```{r ttexample_image, echo=FALSE}
#| label: "fig-tt_healthcare"
#| fig-cap: "Existing surface of motorised travel time from healthcare facilities, exerpted to Cambodia (Weiss et al., 2020)"

@@

Reference documents for `traveltime` are available at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by `@weiss2018global; @weiss2020global`.

@@

Global maps of travel time to cities `[@weiss2018global]` and health care facilities `[@weiss2020global]` `@fig-tt_healthcare` have generated significant interest and use. Collectively `@weiss2018global ; @weiss2020global` have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package `[@geodata]`. There is clear demand for these type of products.

@@

Other R packages produce superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the <https://www.TravelTime.com> API `[@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]`. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `traveltime` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repetitive iteration for all combinations of time and points, and additional calculations. Furthermore, TravelTime.com requires access keys, paid subscription beyond a short free period, caps queries, which add considerable friction to the use experience.

@@

# Example workflow: walking home from public transport in Singapore

@@

## Prepare the data and download friction surface

For this excercise, we need two items of data:

-   our area of interest; in this case, a map of Singapore, and
-   our points to calcute travel time from, that is, the locations of Singapore's MRT and LRT stations.

@@

```{r get basemap}
#| label: "fig-basemap"
#| fig-cap: "Area of interest: Singapore"
library(terra)
library(geodata)
library(tidyterra)

@@

ggplot() +
  geom_spatvector(
    data = sin
  ) +
  # scale_fill_viridis_c(
  #   option = "A",
  #   direction = -1,
  #   na.value = "transparent"
  # ) +
  scale_fill_manual(na.value = "transparent") +
  theme_minimal()
```

@@

We can pass in our basemap `sin` `@fig-basemap`, a `SpatVector`, directly as the `extent`. It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.

@@

```{r  download friction surface, echo=TRUE, output=FALSE}
library(traveltime)

friction_singapore <- get_friction_surface(
    surface = "walk2020",
    extent = sin
  )|> 
  mask(sin)
```

```{r show friction surface raster }
friction_singapore
```

Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points `@fig-data`. Higher values of friction indicate more time travelling across a given cell.

```{r plot data}
#| label: "fig-data"
#| fig-cap: "Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points"

@@

## Calculate and plot the travel time

@@

We present those results in `@fig-result`:

@@

Note that the above raster include infinite (`Inf`) values. In `@fig-data`, a number of islands to the south are shown as cells unconnected with the mainland. These raster cells for these islands appear absent in `@fig-result`. Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot `@fig-result`.
